`make.traits`

### 2023-04-24 inst/dads_paper.Rmd
Update dads_paper.Rmd
Big rename

---
title: "`dads`: a modular `R` package for simulating diversity and traits data."
author: "homas Guillerme, Natalie Cooper, Andrew P. Beckerman, and Gavin H. Thomas"
bibliography: references.bib
csl: mee.csl
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
    keep_tex: true
  self_contained: true
---

@@

<!-- SET URL FOR [manual](dads_manual) -->

@@

3. Here I present the `dads` package, a modular `R` package for diversity and trait disparity simulations.
This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
It also provides a tidy interface through the `dads` object, allowing users to easily run reproducible simulations.

4. The `dads` package also comes with an extend manual regularly updated following users' questions or suggestions.

@@

<!-- % The problems with the absence of modularity -->
Although the packages mentioned above (and more that we forgot to mention/don't know off) and are excellent (i.e. fast, reliable and documented) at the specific tasks they are designed for, they unfortunately don't allow much modifications beyond the parameters implemented for each specific function.
This can be problematic for generating specific simulations to be used as a specific null model for some specific questions.
For example, `TreeSim` can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g. simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
Or `PETER` is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
Although we don't think this was the aim of these specific packages authors (not to create a strawman argument) we think this hampers the use of specific null model (or baseline model) to test specific dynamic mechanisms or to compare to specific datasets. 
<!-- % The solution with dads -->
Therefore, we propose `dads` a modular `R` package to simulate both diversity and disparity through time.

# Description
<!-- % Very briefly how does it work -->
`dads` is based on a core birth-death process implemented in the function `dads` that simulates diversity through time using the following steps:

@@

In `dads`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig.  \ref{Fig_workflow}).
The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.dads` function.

![Figure 1:](gitbook/dads_structure.pdf)

\label{Fig_workflow}`dads` package workflow @@@. `"events"` are an additional category of modular objects discussed later.

@@

```{r, eval = FALSE, echo = TRUE}
## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20))
```

@@

These are included by default in a `dads` process.
The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
Using a reserved syntax in this case the arguments `lineage` and `bd.params` (much more details about the syntax is available in the [manual](dads_manual)). <!-- TODO: make link -->
For example, to simulate the same process but hard coding the modifiers we can use the following functions for each process:

@@

## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      modifiers = my_modifiers)
```

Of course these function can be greatly modified and depend on other parameters of the the simulation.
For example if traits are also simulated (see below) we can bias the branch length to be also dependent on the trait distribution, the selection to be biased towards species at the extremes of the distribution and the speciation modifier to only allow speciation when the lineage's parent trait value is positive.
More example of biased or special `"modifiers"` are available in the [manual](dads_manual).

@@

Another major feature of `dads` is to allow to specify the trait simulation process manually as a `"traits"` object.
This can be done in the same way as for the `"modifiers"` but with the `make.traits` function.
Here the Brownian motion syntax is also specific (especially `x0` and `edge.length`), more details are available in the [manual](dads_manual).

@@

## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
## and a Brownian Motion trait process simulated
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      modifiers = my_modifiers,
                      trait     = my_trait)
```

@@

```{r, eval = TRUE, echo = FALSE}
## Simulating the tree above with a nice looking random seed
## (all the options described above, modifiers and traits are default)
## so it's pretty easy (we don't need to specify the trait process and the modifiers)
set.seed(2)
my_simulation <- dads(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      trait     = make.traits())
# plot(my_simulation)
```

The resulting object is a `"dads"` object containing a `$tree` element (a `"phylo"`) object and a `$data` object containing a `"matrix"` object:

```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
## Printing the dads object
my_simulation

@@

The package also comes with it's own plotting function allowing to directly display `dads` generated tree and data in the same plot:

```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
## A default dads plot
plot(my_simulation, main = "Default plot")
```

Note that here too many options are available for plotting such objects. More information is again available in the [manual](dads_manual).

@@

## `dads` compared to other packages

@@

# Conclusion
The `dads` is modular and nice to use.

# Package location
The `dads` package is available on the CRAN at \url{https://cran.r-project.org/web/packages/dads/index.html} or on GitHub at \url{https://github.com/TGuillerme/dads} with more associated information.
All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

>>>>

---
title: "`treats`: a modular `R` package for simulating diversity and traits data."
author: "homas Guillerme, Natalie Cooper, Andrew P. Beckerman, and Gavin H. Thomas"
bibliography: references.bib
csl: mee.csl
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
    keep_tex: true
  self_contained: true
---

@@

<!-- SET URL FOR [manual](treats_manual) -->

@@

3. Here I present the `treats` package, a modular `R` package for diversity and trait disparity simulations.
This package is based on a standard birth death algorithm that can be easily tuned to the user specific needs by designing their specific trait process, birth-death modifiers and events.
It also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.

4. The `treats` package also comes with an extend manual regularly updated following users' questions or suggestions.

@@

<!-- % The problems with the absence of modularity -->
Although the packages mentioned above (and more that we forgot to mention/don't know off) and are excellent (i.e. fast, reliable and documented) at the specific tasks they are designed for, they unfortunately don't allow much modifications beyond the parameters implemented for each specific function.
This can be problematic for generating specific simulations to be used as a specific null model for some specific questions.
For example, `TreeSim` can simulate a birth-death tree with some extinction event but is not designed to simulated one with an extinction event that leads to the birth-death process to be not diversity dependent anymore (e.g. simulating a release in selection pressure after the extinction process that leads to a different process dominating speciation).
Or `PETER` is not designed to simulate a complex set of traits (say three correlated BM traits and two independent OU ones).
Although we don't think this was the aim of these specific packages authors (not to create a strawman argument) we think this hampers the use of specific null model (or baseline model) to test specific dynamic mechanisms or to compare to specific datasets. 
<!-- % The solution with treats -->
Therefore, we propose `treats` a modular `R` package to simulate both diversity and disparity through time.

# Description
<!-- % Very briefly how does it work -->
`treats` is based on a core birth-death process implemented in the function `treats` that simulates diversity through time using the following steps:

@@

In `treats`, these three or four steps are implemented as modular functions that the user can easily change using an internal class of objects called `"modifiers"` or `"traits"` (Fig.  \ref{Fig_workflow}).
The simulation then outputs a tree (of class `"phylo"` and a associated table of traits - `"matrix"`) that can be visualised using the `plot.treats` function.

![Figure 1:](gitbook/treats_structure.pdf)

\label{Fig_workflow}`treats` package workflow @@@. `"events"` are an additional category of modular objects discussed later.

@@

```{r, eval = FALSE, echo = TRUE}
## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa
my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20))
```

@@

These are included by default in a `treats` process.
The modularity of the packages works with the ability to code the modifiers manually rather than using default ones.
Using a reserved syntax in this case the arguments `lineage` and `bd.params` (much more details about the syntax is available in the [manual](treats_manual)). <!-- TODO: make link -->
For example, to simulate the same process but hard coding the modifiers we can use the following functions for each process:

@@

## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      modifiers = my_modifiers)
```

Of course these function can be greatly modified and depend on other parameters of the the simulation.
For example if traits are also simulated (see below) we can bias the branch length to be also dependent on the trait distribution, the selection to be biased towards species at the extremes of the distribution and the speciation modifier to only allow speciation when the lineage's parent trait value is positive.
More example of biased or special `"modifiers"` are available in the [manual](treats_manual).

@@

Another major feature of `treats` is to allow to specify the trait simulation process manually as a `"traits"` object.
This can be done in the same way as for the `"modifiers"` but with the `make.traits` function.
Here the Brownian motion syntax is also specific (especially `x0` and `edge.length`), more details are available in the [manual](treats_manual).

@@

## Running a simulation with a speciation rate of 1, an extinction rate of 0.1
## and stopping when it reaches 20 taxa with the default birth-death modifiers specified
## and a Brownian Motion trait process simulated
my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      modifiers = my_modifiers,
                      trait     = my_trait)
```

@@

```{r, eval = TRUE, echo = FALSE}
## Simulating the tree above with a nice looking random seed
## (all the options described above, modifiers and traits are default)
## so it's pretty easy (we don't need to specify the trait process and the modifiers)
set.seed(2)
my_simulation <- treats(bd.params = list(speciation = 1, extinction = 0.1),
                      stop.rule = list(max.taxa = 20),
                      trait     = make.traits())
# plot(my_simulation)
```

The resulting object is a `"treats"` object containing a `$tree` element (a `"phylo"`) object and a `$data` object containing a `"matrix"` object:

```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
## Printing the treats object
my_simulation

@@

The package also comes with it's own plotting function allowing to directly display `treats` generated tree and data in the same plot:

```{r, eval = TRUE, echo = TRUE, fig.height = 4, fig.width = 4}
## A default treats plot
plot(my_simulation, main = "Default plot")
```

Note that here too many options are available for plotting such objects. More information is again available in the [manual](treats_manual).

@@

Every bit can be coded as a function.
Many already implemented functions (e.g. in events.modifications) but many more possibilities to modify the current ones or design completely new ones.

## `treats` compared to other packages

@@

# Conclusion
The `treats` is modular and nice to use.

# Package location
The `treats` package is available on the CRAN at \url{https://cran.r-project.org/web/packages/treats/index.html} or on GitHub at \url{https://github.com/TGuillerme/treats} with more associated information.
All the versions of the package are archived on ZENODO with associated DOI \url{https://zenodo.org/@@@}.

### 2023-04-24 inst/gitbook/01_getting_started.Rmd
Big rename

```{r, echo = FALSE}
library(dads)
set.seed(4242)
```

@@

One of the simplest things to do with the `dads` package is to simulate a birth-death tree.
For that you can use the function `dads` and specify your stopping rule.
The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:

@@

```{r}
## Running the birth-death simulation
my_tree <- dads(stop.rule = my_stop_rule)
```

@@

People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
By default `dads` runs a pure birth model (speciation is set to 1 and extinction to 0).
However, you can easily change that by specifying your own birth-death parameters:

@@

```{r}
## Generating a birth-death tree with extinctions:
my_tree <- dads(bd.params = my_params, stop.rule = my_stop_rule)
## Visualising the new tree
plot(my_tree)
```

@@

Chances are that you are using `dads` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
Simulating traits is not much more complicated in `dads`: you'll simply need to create a `"traits"` object using the `make.traits` function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

@@

By default, this trait is called "A".
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `dads` function to generate the tree and the traits:

```{r}
## Simulate disparity and diversity
my_data <- dads(bd.params = my_params,
                stop.rule = my_stop_rule,
                traits    = my_trait)
```

@@

This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `dads` function:

@@

## Simulate disparity and diversity
default_data <- dads(bd.params = extinction_02,
                     stop.rule = living_20,
                     traits    = BM_trait,
                     modifiers = default_modifiers)
default_data
```

@@

First we need to create a `condition` function to trigger this (non) speciation event.
We can do that by specifying our `condition` function (here the `staying.alive` function) to apply our modification.
For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `dads`' understanding of what you want to assess:

@@

We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
If you are an advances `dads` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `dads` called `speciation`.

@@

Note that the `make.modifiers` function tests whether the input is compatible with `dads` by default so unless you have an error message, your `modifiers` will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:

```{r}
set.seed(1)
## Simulate disparity and diversity
biased_data <- dads(bd.params = extinction_02,
                    stop.rule = living_20,
                    traits    = BM_trait,
                    modifiers = negatives_extinct)
biased_data
```

>>>>

```{r, echo = FALSE}
library(treats)
set.seed(4242)
```

@@

One of the simplest things to do with the `treats` package is to simulate a birth-death tree.
For that you can use the function `treats` and specify your stopping rule.
The stopping rule simply tells the birth-death process to stop whenever it reaches one of these three conditions:

@@

```{r}
## Running the birth-death simulation
my_tree <- treats(stop.rule = my_stop_rule)
```

@@

People familiar with the [birth-death models](https://lukejharmon.github.io/pcm/chapter10_birthdeath/) might have noticed that we did not specify two important things here: the speciation parameter (sometimes called "lambda" or "birth") and the extinction parameter (sometimes called "mu", "death" or "background extinction").
By default `treats` runs a pure birth model (speciation is set to 1 and extinction to 0).
However, you can easily change that by specifying your own birth-death parameters:

@@

```{r}
## Generating a birth-death tree with extinctions:
my_tree <- treats(bd.params = my_params, stop.rule = my_stop_rule)
## Visualising the new tree
plot(my_tree)
```

@@

Chances are that you are using `treats` because you also want to simulate traits (disparity) along with your diversity (otherwise, we suggest using the [`TreeSim`](https://github.com/tanja819/TreeSim/) package that provides many more birth-death models).
Simulating traits is not much more complicated in `treats`: you'll simply need to create a `"traits"` object using the `make.traits` function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

@@

By default, this trait is called "A".
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its `"traits"` object to our previous `treats` function to generate the tree and the traits:

```{r}
## Simulate disparity and diversity
my_data <- treats(bd.params = my_params,
                stop.rule = my_stop_rule,
                traits    = my_trait)
```

@@

This will not do anything to our simulations compared to the previous trait and tree simulation but we can provide our `"modifiers"` object to the `treats` function:

@@

## Simulate disparity and diversity
default_data <- treats(bd.params = extinction_02,
                     stop.rule = living_20,
                     traits    = BM_trait,
                     modifiers = default_modifiers)
default_data
```

@@

First we need to create a `condition` function to trigger this (non) speciation event.
We can do that by specifying our `condition` function (here the `staying.alive` function) to apply our modification.
For that we can use the `parent.traits` utility function that is optimised for accessing traits in the birth-death process (but you can of course write your own).
This function takes the `trait.values` and `parent.lineage` arguments, two arguments that you can leave named as they are to facilitate `treats`' understanding of what you want to assess:

@@

We can then provide these two functions (the condition `negative.ancestor` and how to modify the speciation event when this condition is met `staying.alive`).
If you are an advances `treats` user, you can design your own `speciation` function but if you just want to use a normal `speciation` function, you can use the default one from `treats` called `speciation`.

@@

Note that the `make.modifiers` function tests whether the input is compatible with `treats` by default so unless you have an error message, your `modifiers` will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor has a negative trait value:

```{r}
set.seed(1)
## Simulate disparity and diversity
biased_data <- treats(bd.params = extinction_02,
                    stop.rule = living_20,
                    traits    = BM_trait,
                    modifiers = negatives_extinct)
biased_data
```

### 2023-04-24 inst/gitbook/02_make.traits.Rmd
Big rename

```{r, echo = FALSE}
library(dads)
set.seed(123)
```

In `dads`, traits are simulating by providing a `traits` object to the `traits` argument.
This object is generated using `make.traits`.

> **What is a trait in biology?** This is one of these epistemological questions like what _is_ a species and is not covered in this package.
For the purpose of _this manual_, a **biological trait** can be any coherent character of any number of dimensions which may or may not be independent of other traits.
Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
For the purpose of the syntax in _this package_ a **trait in `dads`** is a column in the output trait matrix.
For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
This single "beak shape" **biological trait** is then composed of 912 **traits in `dads`** (i.e. a matrix of $(4+300)\times3$ columns).

@@

Trait processes in `dads` are functions that must always take the following arguments by default.

@@

> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `dads` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

@@

### Using a `"process"` in `dads`

You can design your own process as a function, as long as it has a valid syntax.
Alternatively, the `dads` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.

@@

This creates `"dads"` `"traits"` objects that you can print and visualise using the `plot` function:

@@

Note that you can see the multiple options for plotting the trait process by looking at `?plot.dads` manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `dads` package):

```{r}
## What's actually in that object?
print.dads(my_trait_object, all = TRUE)
```

@@

You can visualise them individually using the `trait` argument in `plot.dads`:

@@

### What is a trait in `dads`?

Because it would be impossible to accommodate all definitions of a trait in `dads` we chose an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
In `dads`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `dads`:

@@

You can also add traits to already existing trait objects using the simple `add` option.
This option just takes a `"dads"` `"traits"` object and the additional process(es) will be added to it. 
For example:

@@

```{r}
set.seed(1)
## Generating a pure birth tree with the background trait
tree_bkg <- dads(stop.rule = list(max.taxa = 20),
                 traits = my_trait)
## This tree has many internal singleton nodes
plot(tree_bkg)
```

@@

You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `dads`.
This will apply the `traits` object to all lineages currently alive at the required time steps.
These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.

```{r}
set.seed(123)
## Generating a birth-death tree with a BM trait and saving steps at specific times
tree_steps <- dads(stop.rule  = list(max.time = 3),
                   bd.params  = list(speciation = 1, extinction = 0.1),
                   traits     = make.traits(),
                   save.steps = c(1/3, 1, 2))
## This also creates internal singleton nodes
plot(tree_steps)
abline(v = 3 - c(1/3, 1, 2), lwd = 0.5, col = "grey")
```

## Traits implemented in `dads`

If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `dads`. You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:

@@

**This bit is still in development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `dads`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

@@

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `dads` processes to set all the arguments with default values (just in case).

>>>>

```{r, echo = FALSE}
library(treats)
set.seed(123)
```

In `treats`, traits are simulating by providing a `traits` object to the `traits` argument.
This object is generated using `make.traits`.

> **What is a trait in biology?** This is one of these epistemological questions like what _is_ a species and is not covered in this package.
For the purpose of _this manual_, a **biological trait** can be any coherent character of any number of dimensions which may or may not be independent of other traits.
Basically anything can be a trait if it's defined by the user (e.g. one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
For the purpose of the syntax in _this package_ a **trait in `treats`** is a column in the output trait matrix.
For example, one might be interested in a biological trait that is "beak shape" which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
This single "beak shape" **biological trait** is then composed of 912 **traits in `treats`** (i.e. a matrix of $(4+300)\times3$ columns).

@@

Trait processes in `treats` are functions that must always take the following arguments by default.

@@

> Note that in this function definition the arguments `x0` and `edge.length` have a default value of `0` and `1` respectively. In practice, these arguments are effectively set to the correct values in the `treats` internal function (i.e. whatever `x0` and `edge.length` are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).

@@

### Using a `"process"` in `treats`

You can design your own process as a function, as long as it has a valid syntax.
Alternatively, the `treats` package has several inbuilt processes, namely a multidimensional Brownian motion (`BM.process`) or a multidimensional Ornstein-Uhlenbeck process (`OU.process`).
You can find the list of implemented processes by looking at the `?trait.process` manual page in `R`.

@@

This creates `"treats"` `"traits"` objects that you can print and visualise using the `plot` function:

@@

Note that you can see the multiple options for plotting the trait process by looking at `?plot.treats` manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the `treats` package):

```{r}
## What's actually in that object?
print.treats(my_trait_object, all = TRUE)
```

@@

You can visualise them individually using the `trait` argument in `plot.treats`:

@@

### What is a trait in `treats`?

Because it would be impossible to accommodate all definitions of a trait in `treats` we chose an arbitrary one: a trait is whatever you define as a trait!
A trait can be uni-dimensional as the measurement of a feature of an organism (e.g. leaf surface, femur length, etc.) but can also be a multi-dimensional description of a feature, for example in 3D geometric morphometric context, a trait could be defined as "position of landmark X" (which will be a trait with three dimensions, _x_, _y_ and _z_) or in ecology, the location of a plant can be expressed as latitude and longitude coordinates.
In `treats`, the **process** corresponds to this trait definition (e.g. a process can be of n-dimensions and represents one organisms feature) and the **traits** represents the number of dimensions in total. 
So in the examples above, this is how the following traits are interpreted by `treats`:

@@

You can also add traits to already existing trait objects using the simple `add` option.
This option just takes a `"treats"` `"traits"` object and the additional process(es) will be added to it. 
For example:

@@

```{r}
set.seed(1)
## Generating a pure birth tree with the background trait
tree_bkg <- treats(stop.rule = list(max.taxa = 20),
                 traits = my_trait)
## This tree has many internal singleton nodes
plot(tree_bkg)
```

@@

You can also simulate a tree by generating traits at specific time steps with the `save.steps` option in `treats`.
This will apply the `traits` object to all lineages currently alive at the required time steps.
These time steps can be either regular by providing a single numeric value; e.g. `save.steps = 0.1` will take a snapshot of the trait values every 0.1 units of time, or specific, by providing a specific set of values; e.g. `save.steps = c(1, 1.2, 3)` will take a snapshot of the trait values at the required time steps.

```{r}
set.seed(123)
## Generating a birth-death tree with a BM trait and saving steps at specific times
tree_steps <- treats(stop.rule  = list(max.time = 3),
                   bd.params  = list(speciation = 1, extinction = 0.1),
                   traits     = make.traits(),
                   save.steps = c(1/3, 1, 2))
## This also creates internal singleton nodes
plot(tree_steps)
abline(v = 3 - c(1/3, 1, 2), lwd = 0.5, col = "grey")
```

## Traits implemented in `treats`

If you don't want to design your own trait process, you can use one of the following trait processes that are currently implemented in `treats`. You can find more information about their many options using their specific manuals in R or the generic `?trait.process`:

@@

**This bit is still in development.**
We highly suggest leaving `test = TRUE` so that `make.traits` returns an error if a process or its additional arguments (`process.args`) are not formatted correctly.
`make.traits` will send an error if the trait cannot be directly passed to `treats`.
However, in some specific cases (again, probably mainly for development and debugging) it could be useful to skip the tests using `test = FALSE`.

@@

As detailed above, any process of your own design will work as long as it is a function that takes at least the arguments `x0` and `edge.length`.
You can be imaginative and creative when designing your own process but here are two detailed example functions for a unidimensional Brownian Motion and Ornstein-Uhlenbeck process that you can use for a start (or not).
Remember it is good practice for `treats` processes to set all the arguments with default values (just in case).
